Temporal ordering and registration of imaging data in cross--sectional studies of development

### 2014-03-25 image_analysis_paper/pnas_draft.tex
updated paper

\begin{itemize}
\item Want to study development
\item Often difficult/impossible to obtain longitudinal data
\item Instead, we are presented with cross--sectional data, and we need to order it to look at dynamics
\item This can be done by hand for small data sets (Figure \ref{fig:fish})
\item large data sets-- need automation (Figure \ref{fig:fluorescent_images})
\item we will present mathematical algorithms to register images and temporally order them
\item Use {\em Drosophila} as a case study
\end{itemize}

@@

Often when one collects images in experiments, each image is taken in a different orientation. 
%
Therefore, the first task in image analysis is to align or {\em register} the set of images. 
%
There are many methods for image registration. 
%
``Template--based'' registration \cite{ahuja2007template} is a common technique, where one selects a specific image or {\em template}, and then tries to optimally register each image to that template.
%
However, this method can result in many errors for noisy images, and can be problematic if a ``good'' template image is not known {\em a priori}. 
%
Instead, we will use a technique called angular synchronization\cite{singer2011angular} to register sets of images.
%
Angular synchronization first computes the pairwise alignments for every possible pair of images, effectively using each image as a template for every other image.
%
From this large collection of pairwise alignments, it then tries to find optimal alignment for each image, such that these individual alignments are most consistent (in some metric) with the pairwise alignments.
%
Because we consider {\em all} pairwise alignments, angular synchronization is often much more robust to noise.

We will first discuss angular synchronization {\em only} considering rotations. 
%
We have converted the two--dimensional images shown in Figure \ref{fig:fluorescent_images} into concentration profiles on a ring (we will address the full two--dimensional images in a subsequent section).
%
Figure \ref{subfig:1d_image_example} shows a typical image that we have converted to a one--dimensional concentration profile; the profile is discretized into 100 points. 
%
Rotation of this image corresponds to shifting (with periodic boundary conditions) the one--dimensional concentration profile. 
%
Figure \ref{subfig:1d_unaligned_unordered} shows the concentration profiles for all the images in our data set. 
%
The concentration profiles have been stacked in an array, such that each row corresponds to one image.
%
Figure \ref{subfig:1d_aligned_unordered} shows the concentration profiles, now aligned using angular synchronization;
the peaks are aligned, even though we input no information about the existence or location of these peaks into our alogrithm.
%
We would like to note that, because angular synchronization only uses pairwise rotation information, the optimal rotations are recovered {\em up to} a global rotation, and we are permitted rotate every point by a constant rotation.

Given our set of registered signals, shown in Figure \ref{subfig:1d_aligned_unordered}, we now wish to order them {\em in time} so that we can reconstruct the developmental dynamics.
%
We will use diffusion maps (DMAPS) \cite{coifman2005geometric} to temporally order our data.
%
DMAPS aims to uncover a parameterization of high-dimensional data sampled from a low-dimensional nonlinear manifold.
%
In our analysis, we will assume that our data lie on a one--dimensional curve/ manifold, and that this one dimension is correlated with time. 
%
Therefore, sorting the data by the value of DMAPS coordinate will order the data along this one--dimensional curve, and will therefore temporally order the data.

We demonstrate how we can apply these techniques to our (aligned) one--dimensional concentration profiles. 
%
We use the Euclidean distance between the aligned concentration profiles (the rows of Figure \ref{subfig:1d_aligned_unordered}) as our pairwise distances for DMAPS (in \eqref{eq:dmaps_W}).
%
Figure \ref{subfig:1d_aligned_ordered} shows the data from Figure \ref{subfig:1d_aligned_unordered}, now sorted by the value of the first (non--trivial) DMAPS component, $\phi_2$. 
%
As expected, the intensity of the dpERK peaks grows as a function of time. 
%
Although DMAPS can recover a good ordering for the data, it cannot tell us about the absolute time of the data; we therefore have no information about the speed of this trajectory.
%
Furthermore, the direction of the trajectory is also unknown; determining which ``end'' of the DMAPS trajectory corresponds to the young embryos, and which ``end'' corresponds to the old embryos is a task that needs to be done by hand.

In many cases, we are interested in both registering and ordering our data.
%
%Our proposed alignment methodology, angular synchronization, utilizes the eigendecompostion of the matrix $H \in \mathbb{R}^{md \times md}$, and our proposed ordering algorithm, diffusion maps, utilizes the eigendecomposition of the matrix $A \in \mathbb{R}^{m \times m}$.
%
We can combine angular synchronization (for registration) and diffusion maps (for temporal ordering) into one eigencomputation that allows us to {\em simultaneously} recover the optimal alignments and the temporal ordering.
%
This technique is called vector diffusion maps (VDM) \cite{singer2012vector}.
%
It not only reduces the number of required eigencomputations, but also is much more robust to noise, as mistakes in alignment are also considered when ordering the data.
%
Figure \ref{subfig:1d_aligned_ordered_vdm} shows the result of aligning and ordering the data in Figure \ref{subfig:1d_unaligned_unordered} using vector diffusion maps. 
%
The data has been ordered by $\psi_{1, 3}$, the first (non--trivial) VDM coordinate.

The same methodologies (with small modifications) can be applied to two--dimensional images.
%
Instead of having $100$--dimensional concentration vectors, we now work with images containing $10000$ pixels.
%
The two--dimensional images are registered with respect to all possible translations and rotations. 
%
We used vector diffusion maps to align and order the images shown in Figure~\ref{subfig:images_scrambled}.
%
The results are shown in Figure~\ref{subfig:images_vdm}.
%
The peaks are now aligned, and the intensity of dpERK grows as a function of time, as expected.

@@

\item Techniques for registration and ordering
\item Very general-- requires little knowledge of the types of images
\item Ordering-- different than other studies (TSP, MST, etc.)
\item assumes one--dimensionality of data, and one dimension is correlated with time
\item gives no sense of absolute time (speed) or space (where the ``zero point'' is)
\end{itemize}

@@

Let $x_1, \dots, x_m \in \mathbb{R}^n$ denote the signals that we wish to align with respect to rotations.
%
We assume that each signal $x_i$ is a noisy rotated copy of the underlying signal $x_{true}$ (which we are {\em not} given), such that 
\begin{equation}
x_i = f(x_{true}, \theta_i) + \xi_i
\end{equation}
where the function $f(x_{true}, \theta_i)$ rotates the signal $x_{true}$ by $\theta_i$ degrees, and $\xi_i$ is a noise term (often Gaussian noise). 
%
Our goal is to obtain estimates of $\theta_1, \dots, \theta_m$ so that we can align $x_1, \dots, x_m$. 
%
Up to noise, 
\begin{equation} \label{eq:pairwise_rot}
x_i \approx f(x_j, \theta_i - \theta_j) ;
\end{equation}
 note that \eqref{eq:pairwise_rot} does not require knowledge of $x_{true}$.
%
We can obtain an {\em estimate} of $\theta_i - \theta_j$ by computing the rotation that optimally aligns $x_j$ to $x_i$, 
i.e., $\theta_{ij} \approx \theta_i - \theta_j$, where
%
\begin{equation} \label{eq:opt_angle}
\theta_{ij} = \argmin_{\theta} \|x_i - f(x_j, \theta)\|^2.
\end{equation}
%
For the one--dimensional profiles shown in Figure~\ref{fig:1d_demo}, to compute the optimal angles in \eqref{eq:opt_angle}, we exhaustively search over all 100 possible shifts/rotations of the signals.

@@

Let $d$ denote the dimension of the rotation matrices we are considering (for our example of planar rotations, $R(\theta_{ij}) \in \mathbb{R}^{2 \times 2}$ and $d=2$; we will write our procedure for general $d$ because we will later consider three--dimensional rotations).
%
We construct the matrix $H \in \mathbb{R}^{md \times md}$, where $H$ is an $m \times m$ matrix of $d \times d$ blocks with
\begin{equation} \label{eq:H_to_R}
H_{ij} = R(\theta_{ij})
\end{equation}
%

We note that, under our assumption that $\theta_{ij} \approx \theta_i - \theta_j$, 
\begin{equation} 
H_{ij} = R(\theta_{ij}) \approx R(\theta_i - \theta_j) = R(\theta_i) R(-\theta_j) = R(\theta_i) R(\theta_j)^T,
\end{equation}
 and so
\begin{equation} \label{eq:H_low_rank}
	H \approx 
	\begin{bmatrix}
	R(\theta_1) \\
	R(\theta_2) \\
	\vdots \\
	R(\theta_m)
	\end{bmatrix}
	\begin{bmatrix}
	R(\theta_1)^T R(\theta_2)^T \dots R(\theta_m)^T
	\end{bmatrix}.
\end{equation}
%
From \eqref{eq:H_low_rank}, we can see that we can (approximately) recover $R(\theta_1), R(\theta_2), \dots, R(\theta_m)$ by computing the top block eigenvector of $H$, which we will denote $\hat{R}$.
%
Let $\phi_1, \phi_2, \dots, \phi_{md}$ denote the eigenvectors of $H$, ordered such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_{md}|$, where $\lambda_i$ is the eigenvalue corresponding to $\phi_i$. 
%
Then the top block eigenvector of $H$, $\hat{R}$, can be written as
\begin{equation}
\hat{R} = 
\begin{bmatrix}
\hat{R}_1 \\
\hat{R}_2 \\
\vdots \\
\hat{R}_m
\end{bmatrix} =
\begin{bmatrix}
| & | & & | \\
\phi_1 & \phi_2 & \dots & \phi_d \\
| & | & & | 
\end{bmatrix},
\end{equation}
and $\hat{R}_i \in \mathbb{R}^{d \times d}$ is the estimate for $R(\theta_i)$. 
%
Because $\hat{R}_i$ is not guaranteed to be a rotation matrix, we must project $\hat{R}_i$ onto the closest rotation matrix \cite{...}. 
%
Let $U_i$ and $V_i$ denote the left and right singular vectors, respectively, of $\hat{R}_i$.
%
Then
\begin{equation}
R_{i, est} = U_i V_i^T
\end{equation} 
where $R_{i, est}$ is an orthonormal matrix and is our estimate of $R(\theta_i)$. 
%
Note that $R_{i, est}$ is not guaranteed to be a proper rotation (i.e. $det(R_{i, est} = \pm 1)$; because we are permitted to scale the eigenvectors of a matrix, we adjust the sign of $\\phi_d$ so that (most of) the rotations are proper. 
%
We can now estimate $\theta_{i}$ by inverting \eqref{eq:R_theta}, and then register the signals by rotating signal $i$ by $-\theta_i$.

Furthermore, this formulation also considers {\em higher--order} consistency information. 
%
For example, given our pairwise estimates $R_{ij}$, we know that relationships of the form
\begin{equation} \label{eq:triplet_consistency}
R(\theta_{ik}) R(\theta_{kj}) \approx R(\theta_i) R(\theta_k)^T R(\theta_k) R(\theta_j)^T = R(\theta_i) R(\theta_j)^T
\end{equation}
should also hold.
%
Note that
\begin{equation}
(H^2)_{ij} = \sum_k R(\theta_{ik}) R(\theta_{kj});
\end{equation}
therefore, {\em all} infomation of the form in \eqref{eq:triplet_consistency} is contained in the matrix $H^2$.
%
Because $H$ and $H^2$ have the same eigenvectors, our problem formulation accounts for not only pairwise alignment information, but also these higher--order considerations.

@@

We will uncover a parameterization of this curve by only considering {\em local} neighbors of each data point.
%
To illustrate how local neighbors can allow us to uncover a global parameterization, assume we have data $x_1, x_2, \dots, x_m \in \mathbb{R}$ such that $x_1 < x_2 < \dots < x_m$.
%
We consider $x_i$ ``close to'' $x_{i+1}$ and $x_{i-1}$, and $x_i$ ``far away from'' all other data points. 
%
We then construct the (symmetric) weight matrix, $W \in \mathbb{R}^{m \times m}$,
\begin{equation}
W = 
\begin{bmatrix}
	1 & 1/2 & 0 & 0 & \dots & 0 \\
	1/2 & 1 & 1/2 & 0 & \dots & 0 \\
	\ddots & \ddots & \ddots & \ddots & \ddots & \ddots \\
	0 & 0 & 0 & \dots & 1 & 1/2 \\
	0 & 0 & 0 & \dots & 1/2 & 1 
\end{bmatrix}.
\end{equation}
%
$W_{ij}$ is large if $x_i$ is close to $x_j$,
%
If we normalize this matrix such that $\sum_j W_{ij} = 1$, then one can show that the first eigenvector of this normalized matrix is the all--ones vector, $\phi_1 = [1 1 \dots 1]^T$, and the second eigenvector is a cosine mode, $\phi_2 = [cos(\pi/m) \cos(2 \pi/ m) \cdots \cos((m-1) \pi / m) \cos(\pi)]^T$.
%
Therefore, the second eigenvector, $\phi_2$ is {\em one--to--one} with the parameterization of the data, i.e. $\phi_2(1) > \phi_2(2) > \dots > \phi_2(m)$, just as $x_1 < x_2 < \dots < x_m$. 
%
Therefore, sorting our data by the value of the entries in $\phi_2$ will allow us to order our data.

For the above analysis, the key step was having a notion of ``closeness'' between data points so that we can construct the matrix $W$.
%
For general data in high--dimensional space, we typically use a Gaussian kernel to construct $W$,
\begin{equation} \label{eq:dmaps_W}
W_{ij} = \exp \left( -\frac{d^2(x_i, x_j)}{\epsilon^2} \right)
\end{equation}
and $\epsilon$ is a characteristic distance between data points.
%
Therefore, points closer than $\epsilon$ apart are considered ``close'' and points farther than $\epsilon$ apart are considered ``far away''.
%
$\epsilon$ can be chosen using several techniques (see, for example \cite{coifman2008graph}); in practice, we choose $\epsilon$ to be the median of the pairwise distances between data points.

We then compute the diagonal matrix $D$, where $D_{ii} = \sum_{j=1}^{m} W_{ij}$, and the matrix $A$, where
\begin{equation} \label{eq:dmaps_A}
A = D^{-1} W.
\end{equation} 
%
We calculate the eigenvectors $\phi_1, \phi_2, \dots, \phi_m$ and eigenvalues $\lambda_1, \lambda_2, \dots, \lambda_m$ and order them such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_m|$ \footnote{Because the matrix $A$ is similar to the symmetric matrix $D^{-1/2} W D^{-1/2}$, $A$ is guaranteed to have real eigenvalues and real, orthogonal eigenvectors.}. 
%
Because the matrix $A$ is row-stochastic, $\lambda_1=1$ and $\phi_1$ is a constant vector.
%
In general, the next few eigenvectors $\phi_2, \dots, \phi_m$ give ``meaningful'' embedding coordinates for the data, such that $\phi_j(i)$ gives the $j^{th}$ embedding coordinate of the $i^{th}$ data point. 
%
In our setup, we assume that our data are inherently one--dimensional, and that this one dimension is correlated with time.
%
Therefore, ordering our data by $\phi_2(j)$ will allow us to automatically order our data in time.

@@

\begin{acknowledgments}
-- text of acknowledgments here, including grant info --
\end{acknowledgments}

@@

%\bibitem{ahuja2007template}
%Sunil Ahuja, Ioannis~G Kevrekidis, and Clarence~W Rowley.
%\newblock Template-based stabilization of relative equilibria in systems with
%  continuous symmetry.
%\newblock {\em Journal of Nonlinear Science}, 17(2):109--143, 2007.

>>>>

\dropcap{E}xperimental studies of developmental dynamics fall in two broadly defined categories: longitudinal and cross-sectional.  
%
In longitudinal studies, a chemical or morphological process is monitored over time within the same embryo.  
%
In a cross-sectional study, one embryo provides just a single snapshot of a population-averaged developmental trajectory which must be reconstructed from multiple snapshots provided by different embryos.
%
Both experimental designs have their advantages and limitations and are extensively used in studies of embryonic development across species. 
%
Our work focuses on cross-sectional studies, which have a time-honored history and still present the only option for most of the organisms.  
%
Most commonly, a group of developing embryos is fixed (arresting their development), stained with chemicals that visualize a handful of cellular processes, and imaged by fluorescence microscopy. 
%
Recent advances in large scale physical manipulation of embryos produce rapidly increasing volumes of imaging data from such experiments, in which every embryo is observed at a different orientation and at a different point in its developmental trajectory. 
%
In order to mine such datasets for underlying developmental dynamics, multiple images must be registered (aligned) and ordered in time. 
%
Here present an algorithmic framework that greatly accelerates both of these tasks.

Temporal ordering and registration of images is straightforward, when the number of images is small and differences between them are apparent. 
%
As an example, Figure~\ref{fig:fish} shows a schematic of fish development, combining the processes of growth and patterning.  
%
In this case, temporal ordering can be accomplished by arranging the fish by their size, which correlates with the progress of development, and image alignment is based on obvious morphological landmarks, such positions of head and fins. 
%
Dealing with real data is far more nontrivial. 
%
To illustrate most of the issues associated with temporal ordering and registration of images, we use a data set from an imaging study of pattern formation in the early {\em Drosophila} embryo, one of the leading systems for studies of developmental dynamics. 
%
This dataset comprises 81 fluorescence microscopy images of optical cross-sections of different embryos fixed during the 3$^{rd}$ hour of their development.  
%
During this time interval, most of cell nuclei in the embryo are arranged under the common plasma membrane and embryo undergoes the process of cellularization, whereby membranes that enclose individual nuclei into different cells grow in length (Figure~\ref{subfig:images_scrambled}).

Each of the images reveals a uniform arrangement of nuclei and nonuniform distribution of two proteins (green and red) involved in patterning of the dorsoventral (back-to-belly) patterning of the embryo. 
%
Each image comes from a different embryo in captured at a different orientation and at a different and unknown point in time (Figure~\ref{subfig:images_scrambled}). 
%
The task is to align these images in a globally consistent way and order them in a way that is most representative of the underlying developmental trajectory. 
%
For this particular dataset, both of these tasks are simplified by prior knowledge. 
%
For instance, it is known that the peak in the distribution of one of the proteins (visualized by the green channel) specifies the ventralmost position in the embryo. 
%
Based on this, images translated to make the centers of the embryos coincide and rotated to place the peak of the green channel at the same angular position. 
%
At the same time, temporal ordering can be done based on the monotonic progress of cellularization. 
%
Figure~\ref{subfig:images_hand_ordered} shows the collection of registered and temporally ordered images.

Note that both of these tasks have been accomplished based on prior knowledge about the underlying developmental processes: We knew that the peak of the green signal can be used to align the images and that the progress of cellularization can order them in time. 
%
Is this information essential? 
%
Below we show that this is not the case and that both the temporal ordering and registration of images can be fully automated, without prior information about the underlying dynamics. 
%
Our algorithmic approach is based on diffusion maps, a broadly applicable technique for the discovery of low-dimensional structures in high-dimensional data sets. 
%
The paper is organized as follows. 
%
In the next two sections, we use a simpler version of the data (one-dimensional spatial profiles) to formulate both tasks mathematically.  
%
Then, we show that both of the tasks can be effectively accomplished in single step. 
%
Following this, we return to the original set of two-dimensional images and demonstrate that our approach can readily align the images and order them correctly in time. 
%
We conclude by discussion of the general requirements that must be satisfied for our approach to work and compare it to other ideas in cross-sectional studies of biological dynamics.

@@

\subsection{Angular synchronization for image registration}

Often when one collects images in experiments, each image is taken in a different orientation. 
%
Therefore, the first task in image analysis is to align or {\em register} the set of images. 
%
This requires translating and rotating each image, so that the set of images are globally aligned.

There are many methods for image registration. 
%
``Template-based'' registration \cite{ahuja2007template} is a common technique, where one selects a specific image or {\em template}, and then translates and rotates each image in the data set until it is optimally aligned with this template (optimal for us means minimizing the norm between the difference in pixel intensities).
%
However, this method can result in many errors if the images are noisy, and can be problematic if a ``good'' template image is not known {\em a priori}. 
%
Instead, we will use a relatively new technique called angular synchronization\cite{singer2011angular} to register sets of images.
%
Angular synchronization first computes the pairwise alignments (translation + rotation) for every possible pair of images, effectively using each image as a template for every other image.
%
From this large collection of pairwise alignments, it then tries to find the optimal translation and rotation for each image, such that these individual alignments are most consistent (in some metric) with the pairwise alignments.
%
The algorithm solves this optimization problem by relaxing it to an eigenproblem; the top eigenvector(s) of a matrix (which contains information about the pairwise alignments) give the translation and rotation for each image.
%
Because the method considers {\em all} pairwise alignments, angular synchronization is often much more robust to noise.

\subsection{Diffusion maps for temporal ordering}

We will use diffusion maps (DMAPS) \cite{coifman2005geometric} to temporally order our data.
%
We assume that our data are high-dimensional (for example, we will work with images of 10000 pixels), but that they lie on a one-dimensional, perhaps nonlinear curve.
%
We further assume that curve can be parameterized by time.
%
Therefore, our goal is to {\em automatically} uncover a parameterization of this curve, so that we can sort/order our data along it, effectively ordering our data in time.

To ``unravel'' this curve, we assume that nearby points (in the high-dimensional space) are also close on the curve.
%
We first construct the matrix $W$ (see Equation \eqref{eq:dmaps_W}); $W_{ij}$ is large if $i$ and $j$ are ``close''.
%
We then want to find a coordinate $y$ such that $y(i)$ and $y(j)$ if $W_{ij}$ is large (i.e., $y$ preserves local information, and so points that are close in the original space are also close in the coordinate $y$).
%
We therefore want to solve the following optimization problem
\begin{equation} \label{eq:dmaps_opt_problem}
\argmin_{y} \sum_{ij} W_{ij} (y(i) - y(j))
\end{equation}
%
The solution \ref{eq:dmaps_opt_problem} is $\phi_2$, the second eigenvector of $A$, where $A = D^{-1} W$, and $D$ is a diagonal matrix with $D_{ii} = \sum_j W_{ij}$ \cite{...}.
%
Therefore, $\phi_2$ provides a one-dimensional parameterization of the data, and ordering the data points by the value of $\phi_2$ will order them in time.

\subsection{Vector diffusion maps for registration and ordering}

In many cases, we are interested in both registering and ordering our data.
%
%Our proposed alignment methodology, angular synchronization, utilizes the eigendecompostion of the matrix $H \in \mathbb{R}^{md \times md}$, and our proposed ordering algorithm, diffusion maps, utilizes the eigendecomposition of the matrix $A \in \mathbb{R}^{m \times m}$.
%
We can combine angular synchronization (for registration) and diffusion maps (for temporal ordering) into one eigencomputation that allows us to {\em simultaneously} recover the optimal alignments and the temporal ordering.
%
This technique is called vector diffusion maps (VDM) \cite{singer2012vector}.
%
It not only reduces the number of required eigencomputations, but also is much more robust to noise, as mistakes in alignment are also considered when ordering the data.

\subsection{Analysis on one-dimensional concentration profiles}

We will first demonstrate our techniques considering {\em only} considering rotations. 
%
We have converted the two-dimensional images shown in Figure \ref{fig:fluorescent_images} into concentration profiles on a ring (we will address the full two-dimensional images in a subsequent section).
%
Figure \ref{subfig:1d_image_example} shows a typical image that we have converted to a one-dimensional concentration profile; the profile is discretized into 100 points. 
%
Rotation of this image corresponds to shifting (with periodic boundary conditions) the one--dimensional concentration profile. 
%
Figure \ref{subfig:1d_unaligned_unordered} shows the concentration profiles for all the images in our data set. 
%
The concentration profiles have been stacked in an array, such that each row corresponds to one image.

Figure \ref{subfig:1d_aligned_unordered} shows the concentration profiles, now aligned using angular synchronization;
the peaks are aligned, even though we input no information about the existence or location of these peaks into our algorithm.
%
We would like to note that, because angular synchronization only uses pairwise rotation information, the optimal rotations are recovered {\em up to} a global rotation, and we are permitted rotate every point by a constant rotation.

Given our set of registered signals, shown in Figure \ref{subfig:1d_aligned_unordered}, we now wish to order them {\em in time} so that we can reconstruct the developmental dynamics.
%
We use the Euclidean distance between the aligned concentration profiles (the rows of Figure \ref{subfig:1d_aligned_unordered}) as our pairwise distances for DMAPS (in \eqref{eq:dmaps_W}).
%
Figure \ref{subfig:1d_aligned_ordered} shows the data from Figure \ref{subfig:1d_aligned_unordered}, now sorted by the value of the first (non-trivial) DMAPS component, $\phi_2$. 
%
As expected, the intensity of the dpERK peaks grows as a function of time. 
%
Although DMAPS can recover a good ordering for the data, it cannot tell us about the absolute time of the data; we therefore have no information about the speed of this trajectory.
%
Furthermore, the direction of the trajectory is also unknown; determining which ``end'' of the DMAPS trajectory corresponds to the young embryos, and which ``end'' corresponds to the old embryos is a task that needs to be done by hand.

Figure \ref{subfig:1d_aligned_ordered_vdm} shows the result of aligning and ordering the data in Figure \ref{subfig:1d_unaligned_unordered} using vector diffusion maps. 
%
The data has been ordered by $\psi_{1, 3}$, the first (non-trivial) VDM coordinate.

\subsection{Analysis of two-dimensional images}

The same methodologies (with small modifications) can be applied to two-dimensional images.
%
Instead of having $100$-dimensional concentration vectors, we now work with images containing $10000$ pixels.
%
The two-dimensional images are registered with respect to all possible translations and rotations; 
we convert two-dimensional translations and rotations to three-dimensional rotations so that we can apply angular synchronization or vector diffusion maps. 
%
We used vector diffusion maps to align and order the images shown in Figure~\ref{subfig:images_scrambled}.
%
The results are shown in Figure~\ref{subfig:images_vdm}.
%
The peaks are now aligned, and the intensity of dpERK grows as a function of time, as expected.

@@

\item Techniques for registration and ordering
\item Very general- requires little knowledge of the types of images
\item Ordering- different than other studies (TSP, MST, etc.)
\item assumes one-dimensionality of data, and one dimension is correlated with time
\item gives no sense of absolute time (speed) or space (where the ``zero point'' is)
\end{itemize}

@@

Let $x_1, \dots, x_m \in \mathbb{R}^n$ denote the signals that we wish to align with respect to rotations.
%
We assume that each signal $x_i$ is a noisy rotated copy of the underlying signal $x_{true}$ (which we are {\em not} given), such that 
\begin{equation}
x_i = f(x_{true}, \theta_i) + \xi_i
\end{equation}
where the function $f(x_{true}, \theta_i)$ rotates the signal $x_{true}$ by $\theta_i$ degrees, and $\xi_i$ is a noise term (often Gaussian noise). 
%
Our goal is to obtain estimates of $\theta_1, \dots, \theta_m$ so that we can align $x_1, \dots, x_m$. 
%
Up to noise, 
\begin{equation} \label{eq:pairwise_rot}
x_i \approx f(x_j, \theta_i - \theta_j) ;
\end{equation}
 note that \eqref{eq:pairwise_rot} does not require knowledge of $x_{true}$.
%
We can obtain an {\em estimate} of $\theta_i - \theta_j$ by computing the rotation that optimally aligns $x_j$ to $x_i$, 
i.e., $\theta_{ij} \approx \theta_i - \theta_j$, where
%
\begin{equation} \label{eq:opt_angle}
\theta_{ij} = \argmin_{\theta} \|x_i - f(x_j, \theta)\|^2.
\end{equation}
%
For the one-dimensional profiles shown in Figure~\ref{fig:1d_demo}, to compute the optimal angles in \eqref{eq:opt_angle}, we exhaustively search over all 100 possible shifts/rotations of the signals.

@@

Let $d$ denote the dimension of the rotation matrices we are considering (for our example of planar rotations, $R(\theta_{ij}) \in \mathbb{R}^{2 \times 2}$ and $d=2$; we will write our procedure for general $d$ because we will later consider three-dimensional rotations).
%
We construct the matrix $H \in \mathbb{R}^{md \times md}$, where $H$ is an $m \times m$ matrix of $d \times d$ blocks with
\begin{equation} \label{eq:H_to_R}
H_{ij} = R(\theta_{ij})
\end{equation}
%

We note that, under our assumption that $\theta_{ij} \approx \theta_i - \theta_j$, 
\begin{equation} 
H_{ij} = R(\theta_{ij}) \approx R(\theta_i - \theta_j) = R(\theta_i) R(-\theta_j) = R(\theta_i) R(\theta_j)^T,
\end{equation}
 and so
\begin{equation} \label{eq:H_low_rank}
	H \approx 
	\begin{bmatrix}
	R(\theta_1) \\
	R(\theta_2) \\
	\vdots \\
	R(\theta_m)
	\end{bmatrix}
	\begin{bmatrix}
	R(\theta_1)^T R(\theta_2)^T \dots R(\theta_m)^T
	\end{bmatrix}.
\end{equation}
%
From \eqref{eq:H_low_rank}, we can see that we can (approximately) recover $R(\theta_1), R(\theta_2), \dots, R(\theta_m)$ by computing the top block eigenvector of $H$, which we will denote $\hat{R}$.
%
Let $\phi_1, \phi_2, \dots, \phi_{md}$ denote the eigenvectors of $H$, ordered such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_{md}|$, where $\lambda_i$ is the eigenvalue corresponding to $\phi_i$. 
%
Then the top block eigenvector of $H$, $\hat{R}$, can be written as
\begin{equation}
\hat{R} = 
\begin{bmatrix}
\hat{R}_1 \\
\hat{R}_2 \\
\vdots \\
\hat{R}_m
\end{bmatrix} =
\begin{bmatrix}
| & | & & | \\
\phi_1 & \phi_2 & \dots & \phi_d \\
| & | & & | 
\end{bmatrix},
\end{equation}
and $\hat{R}_i \in \mathbb{R}^{d \times d}$ is the estimate for $R(\theta_i)$. 
%
Because $\hat{R}_i$ is not guaranteed to be a rotation matrix, we must project $\hat{R}_i$ onto the closest rotation matrix \cite{...}. 
%
Let $U_i$ and $V_i$ denote the left and right singular vectors, respectively, of $\hat{R}_i$.
%
Then
\begin{equation} \label{eq:R_est}
R_{i, est} = U_i V_i^T
\end{equation} 
where $R_{i, est}$ is an orthonormal matrix and is our estimate of $R(\theta_i)$. 
%
Note that $R_{i, est}$ is not guaranteed to be a proper rotation (i.e. $det(R_{i, est} = \pm 1)$; because we are permitted to scale the eigenvectors of a matrix, we adjust the sign of $\phi_d$ so that (most of) the rotations are proper. 
%
We can now estimate $\theta_{i}$ by inverting \eqref{eq:R_theta}, and then register the signals by rotating signal $i$ by $-\theta_i$.

%Furthermore, this formulation also considers {\em higher-order} consistency information. 
%%
%For example, given our pairwise estimates $R_{ij}$, we know that relationships of the form
%\begin{equation} \label{eq:triplet_consistency}
%R(\theta_{ik}) R(\theta_{kj}) \approx R(\theta_i) R(\theta_k)^T R(\theta_k) R(\theta_j)^T = R(\theta_i) R(\theta_j)^T
%\end{equation}
%should also hold.
%%
%Note that
%\begin{equation}
%(H^2)_{ij} = \sum_k R(\theta_{ik}) R(\theta_{kj});
%\end{equation}
%therefore, {\em all} infomation of the form in \eqref{eq:triplet_consistency} is contained in the matrix $H^2$.
%%
%Because $H$ and $H^2$ have the same eigenvectors, our problem formulation accounts for not only pairwise alignment information, but also these higher-order considerations.

@@

%We will uncover a parameterization of this curve by only considering {\em local} neighbors of each data point.
%%
%To illustrate how local neighbors can allow us to uncover a global parameterization, assume we have data $x_1, x_2, \dots, x_m \in \mathbb{R}$ such that $x_1 < x_2 < \dots < x_m$.
%%
%We consider $x_i$ ``close to'' $x_{i+1}$ and $x_{i-1}$, and $x_i$ ``far away from'' all other data points. 
%%
%We then construct the (symmetric) weight matrix, $W \in \mathbb{R}^{m \times m}$,
%\begin{equation}
%W = 
%\begin{bmatrix}
%	1 & 1/2 & 0 & 0 & \dots & 0 \\
%	1/2 & 1 & 1/2 & 0 & \dots & 0 \\
%	\ddots & \ddots & \ddots & \ddots & \ddots & \ddots \\
%	0 & 0 & 0 & \dots & 1 & 1/2 \\
%	0 & 0 & 0 & \dots & 1/2 & 1 
%\end{bmatrix}.
%\end{equation}
%%
%$W_{ij}$ is large if $x_i$ is close to $x_j$,
%%
%If we normalize this matrix such that $\sum_j W_{ij} = 1$, then one can show that the first eigenvector of this normalized matrix is the all-ones vector, $\phi_1 = [1 1 \dots 1]^T$, and the second eigenvector is a cosine mode, $\phi_2 = [cos(\pi/m) \cos(2 \pi/ m) \cdots \cos((m-1) \pi / m) \cos(\pi)]^T$.
%%
%Therefore, the second eigenvector, $\phi_2$ is {\em one-to-one} with the parameterization of the data, i.e. $\phi_2(1) > \phi_2(2) > \dots > \phi_2(m)$, just as $x_1 < x_2 < \dots < x_m$. 
%%
%Therefore, sorting our data by the value of the entries in $\phi_2$ will allow us to order our data.

For diffusion maps, the first step is to construct a matrix $W \in \mathbb{R}^{m \times m}$ that contains information about which pairs of data points are ``close.''
%For the above analysis, the key step was having a notion of ``closeness'' between data points so that we can construct the matrix $W$.
%
For general data in high-dimensional space, we typically use a Gaussian kernel to construct $W$,
\begin{equation} \label{eq:dmaps_W}
W_{ij} = \exp \left( -\frac{d^2(x_i, x_j)}{\epsilon^2} \right)
\end{equation}
and $\epsilon$ is a characteristic distance between data points.
%
Therefore, points closer than $\epsilon$ apart are considered ``close'' and points farther than $\epsilon$ apart are considered ``far away''.
%
$\epsilon$ can be chosen using several techniques (see, for example \cite{coifman2008graph}); in practice, we choose $\epsilon$ to be the median of the pairwise distances between data points.

We then compute the diagonal matrix $D$, where $D_{ii} = \sum_{j=1}^{m} W_{ij}$, and the matrix $A$, where
\begin{equation} \label{eq:dmaps_A}
A = D^{-1} W.
\end{equation} 
%
We calculate the eigenvectors $\phi_1, \phi_2, \dots, \phi_m$ and eigenvalues $\lambda_1, \lambda_2, \dots, \lambda_m$ and order them such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_m|$ \footnote{Because the matrix $A$ is similar to the symmetric matrix $D^{-1/2} W D^{-1/2}$, $A$ is guaranteed to have real eigenvalues and real, orthogonal eigenvectors.}. 
%
Because the matrix $A$ is row-stochastic, $\lambda_1=1$ and $\phi_1$ is a constant vector.
%
In general, the next few eigenvectors $\phi_2, \dots, \phi_m$ give ``meaningful'' embedding coordinates for the data, such that $\phi_j(i)$ gives the $j^{th}$ embedding coordinate of the $i^{th}$ data point. 
%
In our setup, we assume that our data are inherently one-dimensional, and that this one dimension is correlated with time.
%
Therefore, ordering our data by $\phi_2(j)$ will allow us to automatically order our data in time.

@@

\begin{acknowledgments}
- text of acknowledgments here, including grant info -
\end{acknowledgments}

@@

%\bibitem{ahuja2007template}
%Sunil Ahuja, Ioannis~G Kevrekidis, and Clarence~W Rowley.
%\newblock Template-based stabilization of relative equilibria in systems with
%  continuous symmetry.
%\newblock {\em Journal of Nonlinear Science}, 17(2):109-143, 2007.
